---
title: "Exercise-11"
format: html
editor: visual
---

# **Practice ANOVA**

# Preliminaries

-   Using the {tidyverse} `read_csv()` function, load the “AVONETdataset1.csv” dataset from [this URL](https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv) as a “tibble” named **d**. As discussed in class, this is a recently published dataset that compiles morphological measurements and information on various ecological variables and geographic range data for more than 11,000 species of birds.

> -   Winnow the dataset to include only the following variables: **Species1**, **Family1**, **Order1**, **Beak.Length_Culmen**, **Beak.Width**, **Beak.Depth**, **Tarsus.Length**, **Wing.Length**, **Tail.Length**, **Mass**, **Habitat**, **Migration**, **Trophic.Level**, **Trophic.Niche**, **Min.Latitude**, **Max.Latitude**, **Centroid.Latitude**, **Primary.Lifestyle**, \*\* and **Range.Size**
>
> -   Do a bit of exploratory data analysis with this dataset, e.g., using the {skimr} package. Which of the variables are categorical and which are numeric?

```{r,warning=FALSE}
library(tidyverse)
library(skimr)
# Load the dataset
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
d <- read_csv(f, col_names = TRUE)

# Select the 19 specified variables from the dataset
d <- dplyr::select(d, "Species1", "Family1", "Order1", "Beak.Width", "Beak.Depth", "Tarsus.Length", "Wing.Length", "Tail.Length", "Mass", "Habitat", "Migration", "Trophic.Niche", "Min.Latitude", "Trophic.Level", "Max.Latitude", "Centroid.Latitude", "Range.Size", "Beak.Length_Culmen", "Primary.Lifestyle")

# Select quantitative variables
quant_vars <- d %>%
  select_if(is.numeric)

# Select categorical variables
cat_vars <- d %>%
  select_if(is.character)

# Print categorical variables
print(cat_vars)

# Perform exploratory data analysis for quantitative variables
skim(quant_vars)



```

# Challenge 1

### ***One-Factor ANOVA and Inference***

#### **Step 1**

-   Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. lso, you will want to convert the variable **Migration** (which is scored as a number: “1”, “2”, or “3”) from class numeric to either being classified as a factor or as a character (string) variable.A

```{r, warning=FALSE}
library(tidyverse)
library(skimr)

# Read the dataset
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
d <- read_csv(f, col_names = TRUE)

# Filter out missing values
filtered_data <- d %>%
  filter(!is.na(Trophic.Level) & !is.na(Migration) & !is.na(log(Mass)))

# Convert Migration to  character variable
 filtered_data$Migration <- as.character(filtered_data$Migration)  
# Create boxplots
boxplot(log(Mass) ~ Trophic.Level * Migration, data = filtered_data,  
        main = "Boxplots of log(Mass) by Trophic Level and Migration",
        xlab = "Trophic Level", ylab = "log(Mass)")

```

Step 2: a ) Run linear models using the `lm()` function to look at the relationship between log(**Mass**) and **Trophic.Level** and between log(**Mass**) and **Migration**.

```{r, warning=FALSE}
# Run linear regression  model for log(Mass) and Trophic.Level
model_trophic <- lm(log(Mass) ~ Trophic.Level, data = filtered_data)

# Summary of the model
summary(model_trophic)

# Run linear regression model for log(Mass) and Migration
model_migration <- lm(log(Mass) ~ as.factor(Migration), data = filtered_data)

# Summary of the model
summary(model_migration)


```

b ) Examine the output of the resultant linear models. Is log(**Mass**) associated with either **Trophic.Level** or **Migration** category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

1.  For the model examining the relationship between log(mass) and Trophic level, the F-statistic is 78.18 with a very small p-value (\< 0.001), indicating a statistically significant model (F(3, 10982) = 78.18, p \< 0.001). This suggests evidence to reject the null hypothesis of an F value of zero, indicating an association between log(mass) and Trophic level.

2.  Similarly, for the model examining the relationship between log(mass) and Migration, the F-statistic is 144.5 with a very small p-value (\< 0.001). This also indicates that the overall model is statistically significant with (F(2, 10983) = 144.5, p \< 0.001), providing evidence to reject the null hypothesis of a value of zero, indicating an association between log(mass) and Migration

***In conclusion, both Trophic level and Migration are associated with log(mass) based on the F-statistic value.***

c\) Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories.

***Answer:*** Migration categories 2 and 3 differ significantly from the reference level, which is Migration category 1. This is evident from the significant coefficients for Migration levels 2 and 3 compared to the reference level in the above output (p \< 0.001).

To assess the differences between the remaining pair of Migration categories (2 and 3), we can use releveling.

For releveling, we can change the reference level to one of the other categories and compare it against the new reference level. Let's say we want to compare Migration level 1 against Migration level 2, while keeping migration level 3 as the reference.

```{r, warning=FALSE}
# Relevel Migration to compare Migration 1 against Migration 2
filtered_data$Migration <- relevel(factor(filtered_data$Migration), ref = "3")

# Run the linear regression  model 
model_migration_releveled <- lm(log(Mass) ~ as.factor(Migration), data = filtered_data)

# Summary of the re-leveled model
summary(model_migration_releveled)

```

The output indicates that migration categories 1 and 2 are associated with differences in log(Mass) compared to Migration category 3, which is the reference level in this coding. This association is supported by an extremely low p-value (\< 0.001) .

Step 3: Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another

```{r, warning=FALSE}
# Since post-hoc tukey  works with aov function, Fit the model using aov()
model_migration_aov <- aov(log(Mass) ~ as.factor(Migration), data = filtered_data)
summary(model_migration_aov)
# Perform Tukey's HSD test for pairwise comparisons
tukey_result <- TukeyHSD(model_migration_aov)
tukey_result



```

***Interpretation of Post - Hoc Tukey's test*** : The difference in log(Mass) between Migration categories 1 and Migration 3 is estimated to be -0.3765, with very small p value (p = 0), indicating a statistically significant difference between these categories. Similarly, the difference in log(Mass) between Migration categories 2 and Migration 3 is estimated to be 0.3832, with very small p-value (p= 0), indicating a statistically significant difference between these categories. Additionally, the difference in log(Mass) between Migration categories 2 and Migration 1 is estimated to be 0.7597, with very small p value (p = 0), indicating a statistically significant difference between these categories. Overall, these results demonstrate significant differences in log(Mass) among all pairs of Migration categories.

Step 4: Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(Mass) in relation to Trophic.Level and calculate a p value for your original F statistic. You can do this either by programming your own permutation test (e.g., by shuffling values for the predictor or response variable among observations and calculating an F statistic for each replicate) or by using the {infer} workflow and setting calculate(stat="F").

```{r, warning=FALSE}

# Fit the original linear model
original_model <- lm(log(Mass) ~ Trophic.Level, data = filtered_data)

# Calculate the original F statistic
original_anova <- anova(original_model)
original_F_statistic <- original_anova$F[1]

# lets do this using permutation methods
# Number of permutations
n_permutations <- 1000

# Initialize vector to store permuted F statistics
permuted_F_statistics <- numeric(n_permutations)

# Permutation test
set.seed(123) # for reproducibility
for (i in 1:n_permutations) {
  # Permute Trophic.Level
  permuted_data <- filtered_data
  permuted_data$Trophic.Level <- sample(permuted_data$Trophic.Level)
  
  # Fit the permuted linear model
  permuted_model <- lm(log(Mass) ~ Trophic.Level, data = permuted_data)
  
  # Calculate permuted F statistic
  permuted_anova <- anova(permuted_model)
  permuted_F_statistics[i] <- permuted_anova$F[1]
}

# Calculate p-value
p_value <- mean(permuted_F_statistics >= original_F_statistic)
p_value

```

***Interpretation:*** **A p-value of 0 indicates that in the permutation test suggests that the likelihood of observing strong association between log(Mass) and Trophic Level by random chance alone is extremely low. Therefore, we reject the null hypothesis and conclude that there is a significant relationship between log(Mass) and Trophic Level.**

**CHALLANGE 2**

**Step 1: Create the following two new variables and add them to AVONET data-set.**

-   **Relative beak length**, which you should calculate as the *residual* of log(**Beak.Length_Culmen**) on log(**Mass**).

-   **Relative tarsus length**, which you should calculate as the *residual* of log(**Tarsus.Length**) on log(**Mass**).

    ```{r, warning=FALSE}
    library(tidyverse)
    library(dplyr)

    # Fit linear models for log(Beak.Length_Culmen) and log(Tarsus.Length) as functions of log(Mass)
    beak_model <- lm(log(Beak.Length_Culmen) ~ log(Mass), data = d)
    tarsus_model <- lm(log(Tarsus.Length) ~ log(Mass), data = d)

    # Extract residuals
    beak_residuals <- residuals(beak_model)
    tarsus_residuals <- residuals(tarsus_model)

    # Add the residuals as new variables to the AVONET dataset
    d <- d %>%
      mutate(Relative_Beak_Length = beak_residuals,
             Relative_Tarsus_Length = tarsus_residuals)

    d
    ```

#### **Step 2**

-   Make a boxplot or violin plot of your new relative tarsus length variable in relation to **Primary.Lifestyle** and of your new relative beak length variable in relation to **Trophic.Niche.**

```{r, warning=FALSE}
library(ggplot2)

# Remove missing values from Trophic Niche
filtered_data <- d[!is.na(d$Trophic.Niche), ]

# Boxplot of relative tarsus length variable in relation to Primary Lifestyle
ggplot(filtered_data, aes(x = Primary.Lifestyle, y = Relative_Tarsus_Length)) +
  geom_boxplot(fill = "skyblue", color = "blue") +
  labs(x = "Primary Lifestyle", y = "Relative Tarsus Length") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Violin plot of relative beak length variable in relation to Trophic Niche
ggplot(filtered_data, aes(x = Trophic.Niche, y = Relative_Beak_Length)) +
  geom_violin(fill = "lightgreen", color = "darkgreen") +
  labs(x = "Trophic Niche", y = "Relative Beak Length") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

Step 3: a) Run an ANOVA analyses to look at the association between geographic range size and the variable **Migration**. You should first drop those observations for which **Migration** is not scored and also look at the distribution of the variable **Range.Size** to decide whether and how it might need to be transformed. Based on the global model, is range size associated with form of migration? How much of the variance in your measure of range size is associated with **Migration** behavior style?

```{r, eval = FALSE, warning=FALSE}
library(tidyverse)
library(effectsize)

# Drop observations where Migration is not scored
filtered_data <- filtered_data[!is.na(filtered_data$Migration), ]

# Check the distribution of Range.Size
hist(filtered_data$Range.Size, main = "Distribution of Range Size")

# Distribution is positively skewed, lets do log transformation of the data

# Apply log transformation to Range.Size
filtered_data$log_Range_Size <- log(filtered_data$Range.Size)

# Check the distribution of log-transformed Range.Size
hist(filtered_data$log_Range_Size, main = "Distribution of Log-Transformed Range Size")

# Distribution has been approximating normal with log transformation

# Fit the linear regression model (global model)
model <- lm(log(Range.Size) ~ as.factor(Migration), data = filtered_data)
summary(model)

# Calculate effect size (partial eta squared)
partial_eta_squared <- eta_squared(model, partial = TRUE)
partial_eta_squared


```

Yes, range size is significantly associated with the form of migration. This is indicated by the highly significant p-value (\< 0.001) obtained from the ANOVA analysis (F(2, 10929) = 519.5, p \<0.001).The effect size (partial eta squared) for the association between range size and migration behavior style is 0.09. This indicates that approximately 9% of the variance in range size is associated with migration behavior style.

b\) Given the regression coefficients returned in output of the model, which **Migration** categories are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories. Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another

```{r, warning=FALSE}

# In the model provide above, the reference level was 1

# Relevel the Migration variable with migration 2 as the reference level
filtered_data$Migration <- relevel(factor(filtered_data$Migration), ref = "2")

# Fit the linear regression model
model <- lm(log(Mass) ~ Migration, data = filtered_data)
summary(model)

# Conduct post-hoc Tukey HSD test
tukey_result <- TukeyHSD(aov(log(Mass) ~ Migration, data = filtered_data))
tukey_result


```

In the output of the model, Migration2 and Migration3 have coefficients that are different from zero, indicating that these categories are different from the reference level with p value \<0.001.The reference level is Migration1. The data has been re-leveled, with Migration 2 serving as the reference level. The Migration categories 1 and 3 are found to be different from the reference level (Migration category 2) with p value \<0.001. The Tukey HSD test results indicate significant differences in log(Mass) means among all pairs of Migration categories (1-2, 3-2, and 3-1) with p value of 0. Therefore, all Migration categories exhibit statistically significant differences in log(Mass), as indicated by both the regression coefficients and the post-hoc Tukey HSD test.

Step 4: Winnow your original data to just consider birds from the Infraorder “Passeriformes” (song birds).

-   Run separate one-factor ANOVA analyses to look at the association between \[1\] relative beak length and **Primary.Lifestyle** and between \[2\] relative beak length and **Trophic.Level**. In doing so…

    -   Make boxplots of response variable by each predictor and by the combination of predictors.

    -   Run linear models for each predictor separately and interpret the model output.

```{r, warning=FALSE}
library(tidyverse)

# Step 1: Filter original data for birds from the Infraorder "Passeriformes"
passeriformes_data <- filtered_data %>%
  filter(Order1 == "Passeriformes")

# Step 2: Boxplots of response variable by each predictor and by combination of predictors

# Boxplot of relative beak length by Primary Lifestyle
ggplot(passeriformes_data, aes(x = Primary.Lifestyle, y = Relative_Beak_Length)) +
  geom_boxplot() +
  labs(title = "Relative Beak Length by Primary Lifestyle",
       x = "Primary Lifestyle",
       y = "Relative Beak Length")

# Boxplot of relative beak length by Trophic Level
ggplot(passeriformes_data, aes(x = Trophic.Level, y = Relative_Beak_Length)) +
  geom_boxplot() +
  labs(title = "Relative Beak Length by Trophic Level",
       x = "Trophic Level",
       y = "Relative Beak Length")

# Boxplot of relative beak length by combination of Primary Lifestyle and Trophic Level
ggplot(passeriformes_data, aes(x = Primary.Lifestyle, y = Relative_Beak_Length, fill = Trophic.Level)) +
  geom_boxplot(position = "dodge") +
  labs(title = "Relative Beak Length by Primary Lifestyle and Trophic Level",
       x = "Primary Lifestyle",
       y = "Relative Beak Length",
       fill = "Trophic Level")

# Step 3: Run linear models for each predictor separately and interpret the model output

# Linear regression model for Primary Lifestyle
model_primary_lifestyle <- lm(Relative_Beak_Length ~ Primary.Lifestyle, data = passeriformes_data)
summary(model_primary_lifestyle)

# Linear regression model for Trophic Level
model_trophic_level <- lm(Relative_Beak_Length ~ Trophic.Level, data = passeriformes_data)
summary(model_trophic_level)

```

***INTERPRETATION OF MODEL:***

For the primary lifestyle and relative beak length the (F (3, 6602) = 128.9) , p-value (\< 0.001)) indicates that the model is statistically significant, suggesting that there is a relationship between Primary Lifestyle and Relative Beak Length. Similarly, for the trophic level and relative beak length , the (F (2, 6603) = 174.3) , p-value (\< 0.001)) indicates that the model is statistically significant, suggesting that there is a relationship between trophic level and Relative Beak Length.

Step 5:Run a two-factor model to look at the association between relative beak length and both **Primary.Lifestyle** and **Trophic.Level** among the passeriforms. Based on the model output, what would you conclude about how relative beak length is related to these two variables?

```{r, warning=FALSE}
# Run a linear regression model without interaction term
model <- lm(Relative_Beak_Length ~ Primary.Lifestyle + Trophic.Level, data = passeriformes_data)

# View the model summary
summary(model)

```

-   INTERPRETATION: The linear model examining the association between relative beak length and both Primary.Lifestyle and Trophic.Level among the passeriforms reveals several significant findings. Each level of Primary.Lifestyle (Generalist, Insessorial, Terrestrial) demonstrates a significant difference in intercept compared to the reference level, suggesting distinct baseline relative beak lengths for each category (p \< 0.001). Similarly, Trophic.Level Herbivore exhibits a unique intercept, indicating varying baseline relative beak lengths for these categories compared to the reference level (p \< 0.001). However, the coefficient for Trophic.Level Omnivore, does not reach statistical significance with the p value \>0.05. Overall, the model is statistically significant (F = 162.5, p-value \< 0.001), and it explains approximately 10.9% of the variance in relative beak length among passeriforms.

STEP 6:Finally, run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term. To do this, you should modify your model formula using the colon operator (`:`) to specify the interaction, e.g., relative beak length \~ **Primary.Lifestyle** + **Trophic.Level** + **Primary.Lifestyle:Trophic.Level**. Based on the model output, what would you now conclude about how relative beak length is related to these two variables?

```{r, warning=FALSE}
# Run a linear regression model with interaction term 
model <- lm(Relative_Beak_Length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = passeriformes_data)

# View the model summary
summary(model)


```

-   *INTERPRETATION OF RESULT*: Each level of Primary.Lifestyle (Generalist, Insessorial, Terrestrial) is associated with a distinct intercept, indicating the baseline relative beak length for each category compared to the reference level (p \< 0.001). Similarly, each level of Trophic.Level (Herbivore, Omnivore) is associated with a different intercept, signifying the baseline relative beak length for each category compared to the reference level (p \< 0.001).The coefficient of the interaction effect implies that there is no statistically significant impact on relative beak length when both the bird's Primary Lifestyle is Generalist and the Trophic Level is Herbivore (p = 0.0538). However, all other coefficients suggest an additional effect on relative beak length when the bird's Primary Lifestyle is Generalist, Insessorial, or Terrestrial, and its Trophic Level is Omnivore or Herbivore (p \< 0.001).

    The model demonstrates a moderate level of explanatory capability, as indicated by the multiple R-squared value of 13.57%. This suggests that approximately 13.57% of the variance in relative beak length is accounted for by the predictors included in the model. . Furthermore, the F-statistic, with a notably low p-value (\< 0.001), confirms the overall significance of the model.

    Step 7: Use the `interaction.plot()` function to visualize the interaction between **Primary.Lifestyle** and **Trophic.Level**.

```{r, warning=FALSE}
library(ggplot2)

# Create the interaction plot
interaction.plot(
  x.factor = passeriformes_data$Primary.Lifestyle,
  xlab = "Primary Lifestyle",
  trace.factor = passeriformes_data$Trophic.Level,
  trace.label = "Trophic Level",
  response = passeriformes_data$Relative_Beak_Length,
  fun = mean,
  ylab = "Mean Relative Beak Length"
)

```
